---
title: "ezlimma vignette"
author: "Jonathan Dreyfuss"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ezlimma vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction
Welcome to ezlimma. This package is a wrapper for the commonly used limma package, which makes limma easier to use and offers some functionality that adds value to limma's functions, such as mediation analysis and allowing for writing pathway results to Excel workbooks that link to the feature-level (e.g. gene-level) results.

## Install package
See README.md.

## Load package
```{r lib}
library(ezlimma)
```

## Simulate data
To demonstrate the package on a dataset, I'll simulate log2 expression data with 100 genes and 6 samples in two groups, with the first gene up-regulated in the first group. This is the example data from limma `contrasts.fit`.
```{r sim}
set.seed(42)
M <- matrix(rnorm(100*6, sd=0.3), nrow=100, ncol=6)
dimnames(M) <- list(paste0("gene", 1:nrow(M)), paste0("sample", 1:ncol(M)))
grp <- rep(c("First3", "Last3"), each=3)
M[1,1:3] <- M[1,1:3] + 2
```

## Gene level analysis
Assume we want to test if genes are changed in First3, in Last3, or in Last3 - First3. We can do this in two lines of code,
```{r lc}
contr.v <- c(First3="First3", Last3="Last3", Last3vsFirst3="Last3-First3")
res0 <- limma_contrasts(M, grp=grp, contrast.v = contr.v)
```

This returns a data frame that looks like:
```{r, echo=FALSE}
knitr::kable(signif(res0[1:10,], 3))
```

Similarly, we can test association to a phenotype variable. Let's say we have a phenotype vector, 
```{r}
pheno.v <- rnorm(ncol(M))
pheno.v[1:3] <- pheno.v[1:3]-1
```

we can test association with,
```{r}
res1 <- ezcor(M, pheno.v, method="pearson")
```

where `method` can be any of "pearson", "kendall", or "spearman". This returns a data frame like,
```{r, echo=FALSE}
knitr::kable(signif(res1[1:10,], 3))
```

Similarly, we can test association using linear regression with limma, where the model is *gene expression = a + bx + error*, and we test coefficient 2, i.e. the null hypothesis that b=0. So it returns slopes rather than correlation coefficients. The code is,
```{r}
res2 <- limma_cor(M, phenotype = pheno.v)
```

The output is,
```{r, echo=FALSE}
knitr::kable(signif(res2[1:10,], 3))
```

Because this is a regression, we can also adjust for covariates. We do this by creating our own design matrix instead of letting `limma_cor` create one from `phenotype`.
```{r}
covar <- rnorm(ncol(M))
design <- model.matrix(~1+pheno.v+covar)
res3 <- limma_cor(M, design = design)
```

which gives
```{r, echo=FALSE}
knitr::kable(signif(res3[1:10,], 3))
```

We can also test multiple phenotypes together. Say we had a matrix of phenotypes as
```{r}
pheno.mat <- cbind(pheno1=pheno.v, pheno2=rnorm(ncol(M)))
```

then we could test both phenotypes together in one line, as
```{r}
res4 <- multi_cor(M, pheno.mat = pheno.mat, reorder.rows = TRUE)
```

```{r, echo=FALSE}
knitr::kable(signif(res4[1:10,], 3))
```

where the result is ordered by the combined p-value of the two associations per gene. The `method` options for the association test are: "pearson" (the default), "spearman", "kendall", as before, or "limma", where it will use `limma_cor`. This function only allows one method for all phenotypes.

We  can also test what genes mediate the effect of group on the continuous phenotype. This is called high-throughput mediation analysis (`hitman`). We set the exposure (E) to group, the outcome (Y) to `pheno.v`, and test each gene in `M` as a mediator.  
```{r}
res.hit <- hitman(E = grp, M = M, Y = pheno.v)
```

`hitman` returns a table with a combined ("comb") p-value and FDR for each gene. This combined statistic is based on the association of E to the gene in M, and the conditional association of the gene in M to Y given E.
```{r, echo=FALSE}
knitr::kable(signif(res.hit[1:10,], 3))
```

## Gene set level analysis
At a higher level than features (e.g. genes) lie feature sets, such as gene sets or pathways. These can be tested with limma's function `roast`, which are implemented for multiple sets at a time with the functions `mroast` or its faster approximation `fry`. We have written wrappers around these to allow for easier testing, and to make an Excel file where each gene set links to a CSV containing the statistics on its genes, made using the feature-level functions.

Let's make the gene set object,
```{r}
G <- list(list(name="pwy1", description=NA, genes=paste0("gene", 1:10)),
          list(name="pwy2", description=NA, genes=paste0("gene", 11:20)),
          list(name="pwy3", description=NA, genes=paste0("gene", 21:30)))
```
This object would normally be read in using `read_gmt`. Then we can test each of these gene sets.
```{r}
res5 <- roast_contrasts(M, G=G, stats.tab=res0, grp=grp, contrast.v = contr.v, fun="fry", name=NA)
```

If we set `name` to a value other than `NA`, this would create a folder with one CSV per pathway, and an Excel file that looks like
```{r, echo=FALSE}
knitr::kable(res5)
```
except the first column would have links to the CSVs in the folder. Similarly, we can use `roast_cor`.

## System info
Here is the information on the system on which this document was compiled:
```{r si}
sessionInfo()
```